\chapter*{Acknowledgements}
\label{ch:acknowledgements}
\chaptermark{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

The last six years have been exciting, fun, difficult at times, but all in all very rewarding. I feel incredibly lucky to have had people by my side that made every step of the journey so much more fun and easy. I would like to thank them all, because without them, this thesis simply would not have existed. I'm very grateful to every single one of you.

Als eerst wil ik graag alle \textbf{deelnemers aan mijn studie} bedanken, met name alle autistische en sociaal angstige deelnemers. Het is nogal wat om mee te doen met een MRI-experiment of EEG-experiment van een hele dag en op vreemde onderzoekers te moeten vertrouwen dat het allemaal goed gaat. Ik ben dankbaar voor jullie moed en doorzettingsvermogen om mee te doen met deze studie, en ik hoop dat het onderzoek in dit proefschrift een belangrijke bijdrage kan dienen aan het begrijpen van autisme en sociale angst. 

I would like to thank my team of supervisors, whose support and guidance has been essential throughout my PhD trajectory. \newline 
\textbf{Jana}, I don't think I can express my gratitude for you in words, but I will try. Our one-on-one discussions on our results, analyses, and what they might mean, were some of my favorite moments during the past six years. Your fascination with pragmatics is contagious and I will keep it with me for the rest of my life. You kept me motivated during the hard times and I always felt supported and understood by you. Thank you for all the help along on the way. \newline
\textbf{Arjen}, thank you for all the critical questions, feedback and stimulating discussions over the years. Your input always helped elevate the quality of my papers and experiments to a higher level. Your inquisitive attitude is inspiring and I am excited to keep up with your research on mutual understanding in the future. \newline
\textbf{Peter}, your level-headedness and ability to ask exactly the right questions at the right time were very important to the progress in my PhD. I am grateful for your honesty, your focus on finishing the thesis, and your nudges to keep me in the right direction. I feel very lucky to have had you on my team as promotor. 

Between my work and the MPI and the Donders, I have had the chance to work with many wonderful, inspiring and generous colleagues. I would like to thank all of you for making my job so much more fun and my research better. \newline
\textbf{Saskia}, I feel like we worked so closely together and I have learned so much from you, it would be just as fitting to address you in the supervisors section. Statistics, coding in R/Matlab, experiment design, fMRI analysis: you helped me become proficient in all of these with your continuous support and assistance. Even when my motivation was running low, you were there to listen and offer your advice. I could not have wished for a better postdoc to have had alongside me on the Synergy rollercoaster we were on. \newline
\textbf{Ashley}, thank you so much for teaching me all about EEG and signal analysis in the last leg of my PhD. I had a lot of fun learning so much about a topic that was new to me and really enjoyed all of our discussions. I feel like, in your work, you strike the perfect balance between scientific rigor and efficacy and I aspire to do the same in the future. I'm happy to know that many students in Hong Kong will benefit from your excellent supervision. \newline
\textbf{Miriam}, it has been very inspiring to collaborate with you, your endless passion for communication, and your tendency to keep asking questions. We spent so much time coding the behavioral responses of the movie questionnaire, I wonder if anyone in the history of mankind are better experts on what exactly constitutes a mental state word or not. I'm very grateful for the time and effort you spent on our project, and had a lot of fun working on it with you. \newline
\textbf{Maartje}, thank you so, so much for spearheading the participant recruitment and planning of experiment sessions. Our project was so logistically difficult that I am amazed we pulled it off at all and collected all the data that we did. I am so happy and grateful for the effort you put into the Synergy project and to have profited off your patience, decisiveness and the thinking on your feet. \newline
\textbf{Franziska}, it was so fun to be able to chat and collaborate with you on the pupillometry analysis! Thank you for all your preprocessing work on the pupillometry data, the improvements in the analysis pipeline, and the lovely conversations and cups of tea and coffee we shared. \newline
\textbf{Ivan}, thank you for the useful feedback on all of my presentations, posters, and papers. Having an extra professor in my team, and a critical one such as yourself, has been very helpful in improving my work and my thinking. \newline
\textbf{Simon}, for the relatively brief period that you were in Nijmegen, I appreciate the comments and discussions we had during and outside of the Synergy meetings about my research. Thank you for this, as well as being great company during the INSAR conference in Stockholm. \newline
\textbf{Daniela}, I had a great time digging into fMRI analysis with you and thoroughly enjoyed all of our nerdy conversations about movies during the scanning sessions. I promise I will include a note of caution if I recommend you another Austrian thriller. \newline
\textbf{Robin}, thank you for all the work you put in the Synergy project and the Social Anxiety study. I am very proud of the paper and hope the journal we submitted it to will recognize this too. I had a lot of fun working with you on the study and am excited to see you flourish in your career. \newline
Thank you \textbf{Sophia}, \textbf{Christine}, and \textbf{Mirthe} for running the MRI and EEG experiments together. We were working as a well-oiled participant scanning machine by the end and I have all of your problem solving abilities to thank for that. \newline
\textbf{Kexin}, you seemed to put all of your attention and energy in everything you set your mind to. I am very grateful that my research happened to be among those things. From the moment you joined the Synergy lab, I was in awe about how fast you picked up on the details of the study. With the same dedication and insight, you developed an eyetracking analysis pipeline with Arjen, which helped me immensely for my thesis. You were also always on top of the EEG testing sessions and really effortless to collaborate with. I can't imagine this thesis without your help and will always be grateful to have been able to meet you and work with you. \newline
A big thank you to all the \textbf{NBL members} for the insightful presentations, discussions, and your useful feedback when I had a chance to present my work. I'd also like to thank all of the \textbf{NBL PhDs} for your critical feedback and all of you being such nice company. \textbf{Sophie}, thank you for all the productivity meetings and cups of coffee at the height of the pandemic, it helped keep me sane. \textbf{Laura}, thank you for all the advice and insights I needed when finishing up the PhD. A special thanks to rest of the \textbf{NBL Social Committee}, Ashley, Laura, and M\"{u}ge, I really enjoyed the drinks and events we organized together. \newline
An important part in looking forward to going to the office everyday were the lovely people to chat to over lunch at the DCCN. Thank you to \textbf{Kevin}, \textbf{Elena}, \textbf{Egbert}, \textbf{Filip}, \textbf{Helena}, \textbf{Martin}, \textbf{Xiaochen}, \textbf{Anouk}, \textbf{Jorryt}, \textbf{Rene}, \textbf{Felix}, \textbf{Anneloes} and many others for all the good times. \newline
Thank you to all of the PhDs and Postdocs in \textbf{Language in Interaction} for the good company during the retreats. I'd like to thank \textbf{Elena} and \textbf{Marlou} in particular for the useful and fun writing sessions. \newline
To all of my office mates: thank you for both the quick chats and the deep conversations. There have been too many of you too many to name over the course of six years. I hope you all know that you made my day-to-day life a lot more enjoyable. \newline
Thank you \textbf{Annemiek Barsingerhorn} for our conversations and your guidance when I lost track of what was important in the PhD. \newline
Lastly, I want to thank \textbf{Alexandra Elbakyan} for helping researchers around the world like me getting the necessary information to do their work. 

Naast mijn collega's heb ik het geluk om omringd te zijn door veel geweldige mensen die ik met trots mijn vrienden noem. \newline
\textbf{Cansel}, ik ben ontzettend blij en dankbaar dat jij mijn vriendin bent. Je enthousiasme, energie, en je blik op de wereld voelt altijd als een voorrecht om bij in de buurt te zijn. Bedankt voor alle steun en gezelligheid de afgelopen zes jaar. Ik hoop dat er nog heel veel meer komen! \newline 
\textbf{Sophie}, ik ben erg ge\"{i}nspireerd door de manier waarop jij in het leven staat: warm, vriendelijk, kritisch en achtervolgen wat jouw gelukkig maakt in het leven. Ik geniet altijd van onze gesprekken en tijd samen, die me veel steun en inzicht gebracht hebben de afgelopen jaren. \newline 
\textbf{Olaf}, bedankt voor de gezelligheid, het kijken van rare en depressieve films met mij, je eerlijke mening en welkome advies als ik dat nodig heb. Misschien laat ik het wel toe dat we Tusk 2 gaan kijken samen, als die er ooit komt (hopelijk niet). \newline
\textbf{Sjors G.}, bedankt voor de programmeerhulp, relatieadvies, werkplekaanbod en alle andere manieren waarop je me hebt geholpen de afgelopen jaren. Je bent even gul als dat je leuk bent om bij in de buurt te zijn en ik voel me heel dankbaar om jouw als vriend te hebben. \newline 
\textbf{Teun}, ik ben zo blij om je beter te hebben leren kennen vanuit werk! Je bent verantwoordelijk voor heel veel goede feestjes die ik heb mogen bijwonen de laatste jaren en kijk uit naar veel meer nu we praktisch buren zijn! \newline
\textbf{Orhun}, there is little I enjoy more than listening to your rants about the most and very least important things in life, after having watched a three-hour NBC banger of a movie. Thanks for the music recommendations, amazing veggie wraps, for all the Party Den parties and the support when I needed it. \newline 
\textbf{Jonathan}, ik kijk met veel vreugde en dankbaarheid terug op de tijd die we samen besteed hebben. Je oneindige geloof in me heeft me heel veel gesteund toen ik het nodig had. Bedankt voor de fijne tijd en herinneringen. \newline 
Ik geniet van veel dingen in het leven, waaronder films kijken, gamen, muziek luisteren, naar concerten gaan, bordspellen spelen, wandelen, klimmen, en muziekquizzen doen en organiseren. Al deze dingen zijn echter veel leuker om te doen met vrienden dan in je eentje. Heel erg bedankt aan alle mensen die mijn favoriete bezigheden samen met mij hebben willen ondernemen in de afgelopen jaren, waaronder \textbf{Rik}, \textbf{Gerdriaan}, \textbf{Koert}, \textbf{Marion}, \textbf{Louis}, \textbf{Sjors C.}, \textbf{Mike}, \textbf{Amber}, \textbf{Yorick}, \textbf{Annemiek}, \textbf{Judith}, \textbf{Willanka}, \textbf{Yurre}, \textbf{Marlijn}, \textbf{Roel B.}, \textbf{Aaron}, en \textbf{Lenno}. \newline
Bedankt aan alle leuke mensen in \textbf{\#RU}, en in het bijzonder \textbf{\#promovendi}/\textbf{\#academia}, bij wie ik altijd terecht kan voor gezelligheid en steun. Een hechte en ondernemende vriendengroep zoals deze is zeldzaam, en dat koester en waardeer ik enorm.

\textbf{Pap en Mam}, bedankt voor jullie oeverloze steun en vertrouwen tijdens mijn PhD-traject. Jullie zijn er om mij aan te moedigen in alles wat ik doe, en te luisteren en advies te geven als ik dat nodig heb. Jullie geloof en vertrouwen in mij heeft me veel geholpen en ik ben heel dankbaar om jullie altijd achter me te hebben staan. 
\textbf{Laura}, ik prijs me gelukkig met een zus zoals jij. Je bent mijn cheerleader als het goed gaat en een bron van steun als ik met een probleem zit. Je aandacht voor de mensen om je heen is inspirerend, en ik kijk uit naar veel leuke uitjes en successen om met elkaar te vieren. \textbf{Vince}, bedankt voor alle gezelligheid, lol, en gastvrijheid de afgelopen jaren. Het is altijd een plezier om op bezoek te komen en bij te praten in Amsterdam-Noord. 

\textbf{John}, het is altijd leuk met je. Bedankt voor de ontelbare fijne momenten, avonturen, geduld, silliness, vrijheid, en steun die je me gegeven hebt. Ik kijk heel erg uit naar wat de toekomst voor ons samen zal brengen.  