\phantomsection
\addcontentsline{toc}{section}{Research Data Management}
\section*{Research Data Management}
This research followed the applicable laws and ethical guidelines. Research Data Management was conducted according to the FAIR principles. The paragraphs below specify in detail how this was achieved.

\subsection*{Ethics}
This thesis is based on the results of human studies, which were conducted in accordance with the principles of the Declaration of Helsinki. The Ethical Committee of the faculty of Social Sciences (ECSS) has given a positive advice to conduct these studies to the Dean of the Faculty, who formally approved the conduct of these studies (CMO:2019/6059). This research is funded by Gravitation Grant 024.001.006 of the Language in Interaction Consortium from the Netherlands Organization for Scientific Research (NWO).

\subsection*{Findable Accessible}
The table below details where the data and research documentation for each chapter can be found on the Donders Repository (DR). All data archived as a Data Sharing Collection remain available for at least 10 years after termination of the studies.

\begin{table}[ht]
    \captionsetup{justification=raggedright, singlelinecheck=false, font = normal} % Left-align the caption
    \setlength{\tabcolsep}{11pt}
    \renewcommand{\arraystretch}{1.2} % Adjust row spacing (default is 1.0)
    \begin{tabular}{lllll}
    \hline
    \textit{Chapter} & \textit{DAC} & \textit{RDC} & \textit{DSC} & \textit{DSC License} \\
    \hline
    2 & 3025003.02\_119 & 3025003.02\_825 & 3025003.02\_622 & CC0-1.0 \\
    3 & 3025003.02\_119 & 3025003.02\_613 & 3025003.02\_724 & CC0-1.0 \\
    4 & 3011225.01\_121 & 3011225.01\_944 & 3011225.01\_332 & CC0-1.0 \\
    \hline
    \multicolumn{5}{l}{{\fontsize{9.5pt}{0pt}\selectfont DAC = Data Acquisition Collection, RDC = Research Documentation Collection, DSC = Data Sharing Collection}} \\
    \end{tabular}
\end{table}

Informed Consent was obtained through paper forms following the procedure of the Donders Centre for Cognitive Neuroimaging (DCCN). These forms are archived for 10 years in the central archive of the DCCN after termination of the studies. 
    
\subsection*{Interoperable, Reusable}
The raw data is stored on the DAC in the form it was acquired. The data of the DAC of chapter~\ref{ch:mentalizing_asc} and~\ref{ch:mentalizing_sa} and the DSC of chapter~\ref{ch:mentalizing_asc} have been organized according to BIDS standards. For DSC long-lived file formats (.csv, .txt) have been used for most data ensuring that they remain usable in the future. Results are reproducible by providing a description of the experimental setup, raw data (DAC and DSC), analysis scripts or pipelines (RDC and DSC).

\subsection*{Privacy}
The privacy of the participants in this thesis has been warranted using random individual subject codes. A pseudonymization key linked this random code with the personal data. This pseudonymization key was encrypted with a password and stored on a network drive that was only accessible to members of the project who needed access to it because of their role in the project. The pseudonymization key was stored separately from the research data. MRI-data in chapter~\ref{ch:mentalizing_asc} is defaced, not identifiable, and shared without restrictions.

